Galaxy Zoo: Morphological Classifications for 150,000 Galaxies in HST Legacy Imaging

### 2016-05-26 writeup/gz_hubble_data.tex
Significant rewrite of comparisons with volume-limited sample. Closes #133.

Morphologies for GEMS galaxies were measured by \citet{hau07}, who used
single-component \sersic{} fits to the F850LP imaging. A 1\arcsec~positional
match for the GEMS galaxies in Table~9 in \citet{hau07} to the GZH sample gives
8,846~galaxies in both \citet{hau07} and GZH. The primary morphological
parameter in the automated catalog is the \sersic{} index $n$ defining the
radial surface brightness profile.  ``Elliptical'' galaxies are selected by
$n>2.5$ and ``spirals''  by $n>2.5$.  There is no automatic measurement of
irregular or clumpy structure in this catalog.

AEGIS galaxies were morphologically classified using non-parametric
measurements by \citet{lot08}. This method used a combination of the Gini
coefficient ($G$), which measures the relative inequality in pixel brightness,
and $M_{20}$, the second-order moment of the brightest 20\% of the light
\citep{lot04}. A linear combination of $G$ and $M_{20}$ delineates three broad
categories of galaxy morphology: E/S0/Sa (``elliptical''), Sb/Sc/Ir
(``spiral''), and mergers (``irregular''). A 1\arcsec{} positional match on
AEGIS and GZH yields 4,031~galaxies with reliably-measured morphologies and
$S/N>3$ in both $V$- and $I$-bands.

Galaxies in both of the GOODS fields down to a limit of $z_{AB}=22.5$ were
visually classified by a single expert (R.S.~Ellis), inspecting both $z$-band
and composite $Viz$ color images \citep{bun05}. These morphologies are assigned
a numerical value based on categories in \citet{bri98a}: the corresponding
morphologies used are ``elliptical'' (classes 0,1,2), ``spirals'' (classes
3,4,5), and ``irregular'' (classes 6,7,8). A 0.5\arcsec{} matched radius yields
2,435~galaxies (1300 in GOODS-N, 1135 in GOODS-S) in both \citet{bun05} and
GZH.

COSMOS galaxies have multiple published datasets automatically classifying
morphology, all using a variation of non-parametric measurements. \citet{cas07}
used a combination of concentration ($C$), asymmetry ($A$), $G$, and $M_{20}$
\citep{cas05} to classify all galaxies with $m_\mathrm{I,petro}<25$, and used
an empirical division based on several hundred training images to assign
galaxies to discrete morphological categories. A similar method is employed by
\citet{tas11}, using the same non-parametric indices but with a different
method of calculating the Petrosian radius and total light profile. They employ
a nearest-neighbors method to assign morphological categories for the
entire sample based on a small training set. \citet[][ZEST]{sca07} used $C$,
$A$, $G$, $M_{20}$, the galaxy ellipticity ($\epsilon$), and \sersic{} index
($n$) to quantify the galaxy shapes; a principal component analysis is used to
assign galaxies to discrete morphological categories. The categories in all
three methods are split into ellipticals, spirals, and peculiar/irregulars. A
0.5\arcsec{} radius search on GZH matches to 76,241~galaxies in \citet{cas07},
76,608 in \citet{tas11}, and 77,425 in \citet{sca07}.

\begin{figure*}
\center
\includegraphics[width=1.0\textwidth]{figures/comparisons.pdf}
\caption{Distributions of morphological parameters for galaxies matched in GZH
and other published catalogs, split by survey (AEGIS, COSMOS, GEMS, and GOODS).
The first and third columns show the fraction of overall galaxies in the
matched samples, split by their published morphologies as a function of the GZH
\fbest. The corresponding second and fourth columns show data for the same
galaxies as a function of \fodd. The ``odd'' samples exclude galaxies dominated
by lenses or dust lanes ($<1\%$ of the total), which are GZH categories not
expected to correlate to larger-scale structural parameters. Confidence
intervals for each binned fraction in the bin are calculated for a binomial
population \citep{cam11}.}
\label{fig:comparisons}
\end{figure*}

Figure~\ref{fig:comparisons} shows the proportion of galaxies as split by their
automated/expert visual morphologies for each of the six catalogs matched to
GZH. The relation between GZH and the automated/expert morphologies varies
significantly, depending on the survey --- this could be due to intrinsic
differences in the galaxies, in the techniques used to measure morphology, or
both. The distribution of \fbest{} for AEGIS galaxies is essentially flat for
spirals, showing almost no difference in proportion at the highest and lowest
ends. Early-type galaxies are at their highest proportion at \fbest$\sim0$, but
constitute at most only 40\% of galaxies. In contrast, GOODS galaxies at low
\fbest{} are dominated by early-types, although spiral and irregular galaxies
have comparable sizes as low as \fbest$\sim0.2$.

In GEMS galaxies, the impact of irregular or peculiar galaxies is not measured;
splitting on $n=2.5$ gives an essentially flat proportion of ellipticals to
spirals at \fbest$>0.3$.  Visual inspection of images of galaxies where both
\fbest{} and $n$ are high show that the majority are obvious spirals but with
prominent bulges, indicating that the single-component \sersic{} fit is likely
choosing too small of an effective radius and missing the extended disk
structure for a large population of galaxies. This is a known issue with using
$n$ as a standalone measure of galaxy morphology -- it is relatively common,
for example, to measure high $n$ for an object with an exponential disk but a
central cusp in the light profile (eg, due to an AGN, cluster, foreground
star, noise, etc).

Even within the same sample of galaxies in COSMOS, different techniques of
measuring morphology show significant differences. Both \citet{cas07} and
\citet{tas11} have higher proportions of early-type galaxies at \fbest$\sim0$,
but constitute less than half the sample in both cases. The majority of all
galaxies in \citet{sca07}, in contrast, are spirals; oddly, the \emph{highest}
proportions of spirals are found at the lowest values of \fbest. The decrease
in spiral galaxies at higher \fbest{} is almost entirely balanced by an
increase in irregulars. Early-types have a small bump near \fbest$\sim0.15$,
but never constitute more than 10\% of the sample.

The second set of plots in Figure~\ref{fig:comparisons} indicates that \fodd{}
is an effective method for separating irregular and/or peculiar
galaxies from spirals. Despite the somewhat vague wording of the {\it
``anything odd?''} question in GZH, the AEGIS, COSMOS, and GOODS all show a
marked increase in the irregular fraction increasing with \fodd. Any threshold
value for distinguishing between the two, however, depends on
both the survey and redshift~range covered.

As the data shown in Figure~\ref{fig:comparisons} cover a wide range of
redshifts and sizes, a volume-limited sample of galaxies (or at least binning
by redshift) is likely a much more appropriate comparison. While detailed
analysis is left to a further paper, we note that there is no strong change in
the proportion of galaxies when binning by redshift intervals of $\Delta z=0.2$
out to $z=1.0$.

Finally, there are 7,681~galaxies in the GOODS-S field with morphological
classifications in both GZH and the Galaxy~Zoo:~CANDELS project (Simmons
et~al., submitted). Since both the sensitivity and filters for the two sets of
images differ (and there is no correction for redshift bias applied to GZC), there is no
prior reason to expect a perfect correlation between the separate vote
fractions for the projects. Briefly, we note that the \ffeatures{} value for
GZH is on average higher than GZC; the effect is strongest at
$f_\mathrm{features,GZC}<0.3$, for which roughly half the galaxies have
$f_\mathrm{features,GZH}>0.5$. However, the correlation between vote fractions
is single-valued (although not linear, with a Pearson correlation coefficient
of $r=0.6$), and should be possible to calibrate using a similar approach to
that described in Section~\ref{sec:debiasing}; the correlation between other
tasks, such as edge-on galaxies is significantly stronger ($r=0.9$). So while
the raw vote fractions are not directly comparable, the initial analysis
indicates that the broad morphologies are at least consistent.

>>>>

The comparison and analysis of the GZH morphologies is done a set of matched,
volume-limited samples at $z<1.0$ and $m_{I|i|z} < 22.5$~mag.  The redshift and
magnitude limits for the sample are chosen to match the shared constraints for
the shallowest depths \citep[GEMS;][]{bun05} and the limits on morphological
reliability \citep[COSMOS/ZEST;][]{sca07} in the comparison catalogs. We match
all catalogs against GZH using a positional radius of 0.5\arcsec.

Morphologies for GEMS galaxies were measured by \citet{hau07}, who used
single-component \sersic{} fits to the F850LP imaging. A 1\arcsec~positional
match for the GEMS galaxies in Table~9 in \citet{hau07} to the GZH sample gives
8,846~galaxies in both \citet{hau07} and GZH. The primary morphological
parameter in the automated catalog is the \sersic{} index $n$ defining the
radial surface brightness profile.  ``Elliptical'' galaxies are selected by
$n>2.5$ and ``spirals''  by $n<2.5$.  There is no automatic measurement of
irregular or clumpy structure.

AEGIS galaxies were morphologically classified using non-parametric
measurements by \citet{lot08}. Their method used a combination of the Gini
coefficient ($G$), which measures the relative inequality in pixel brightness,
and $M_{20}$, the second-order moment of the brightest 20\% of the light
\citep{lot04}. A linear combination of $G$ and $M_{20}$ delineates three broad
categories of galaxy morphology: E/S0/Sa (``elliptical''), Sb/Sc/Ir
(``spiral''), and mergers (``irregular''). We limit the AEGIS galaxies to those
with reliably-measured morphologies and $S/N>3$ in both $V$- and $I$-bands in
\citet{lot08}.

Galaxies in both of the GOODS fields down to a limit of $z_{AB}=22.5$ were
visually classified by a single expert (R.S.~Ellis), inspecting both $z$-band
and composite $Viz$ color images \citep{bun05}. These morphologies are assigned
a numerical value based on categories in \citet{bri98a}: the corresponding
morphologies used are ``elliptical'' (classes 0,1,2), ``spirals'' (classes
3,4,5), and ``irregular'' (classes 6,7,8).

COSMOS galaxies have multiple published datasets automatically classifying
morphology, all using a variation of non-parametric measurements. \citet{cas07}
used a combination of concentration ($C$), asymmetry ($A$), $G$, and $M_{20}$
\citep{cas05} make empirical divisions based on several hundred training images
to assign galaxy morphology. A similar method was employed by \citet{tas11},
using the same non-parametric indices but with a different method of
calculating the Petrosian radius and total light profile, and using a
nearest-neighbors method to label morphologies. \citet[][ZEST]{sca07} used $C$,
$A$, $G$, $M_{20}$, the galaxy ellipticity ($\epsilon$), and \sersic{} index
($n$) to quantify morphology; a principal component analysis is used to assign
galaxies to categories. All three methods use discrete morphological categories
of ellipticals, spirals, and peculiar/irregulars. We use the same categories
with the exception of assigning S0/bulge-dominated disks (type~2.0) from ZEST
into the elliptical category.

\begin{figure*}
\center
\includegraphics[width=1.0\textwidth]{figures/comparisons_features.pdf}
\caption{Distributions of morphological parameters for a volume-limited sample
($z<1.0$, $m<22.5$) of galaxies matched between GZH and six published
morphological catalogs, split by survey (AEGIS, COSMOS, GEMS, and GOODS). This
plot shows the fraction of overall galaxies for each of the external
morphological categories as a function of \fbest{} in GZH. The shaded regions
around the binned fractions are confidence intervals calculated for a binomial
population \citep{cam11}.}
\label{fig:comparisons_features}
\end{figure*}

Figure~\ref{fig:comparisons_features} shows the proportion of galaxies as split
by their automated/expert visual morphologies for each of the six catalogs
matched to GZH. Galaxies from every catalog show a strong correlation between
\fbest{} and the fraction of galaxies identified as spirals, with a
corresponding anti-correlation between \fbest{} and the fraction of
ellipticals.

The AEGIS sample had the fewest number of matched galaxies (485) with GZH;
while the statistical uncertainties are as a result higher, the strongest
difference between spirals and early-types is seen at \fbest$>0.8$. GOODS
galaxies, which are the only comparison sample that also uses visual
morphology, shows a very similar trend, with nearly 100\% pure samples at
the lowest and highest values of \fbest.  Galaxies from GEMS (which are
only split into two types) also have a strong correlation with \fbest,
which is somewhat surprising (but encouraging) considering the known limits
of assigning galaxy morphology based only on a single \sersic{} fit.

Visual inspection of images of the exceptions --- where both \fbest{} and $n$
are high --- show that most are obvious spirals but with prominent bulges. This
means that the single-component \sersic{} fit is likely choosing too small of
an effective radius and missing the extended disk structure for a large
population of galaxies. This is a known issue with using $n$ as a standalone
measure of galaxy morphology -- it is relatively common, for example, to
measure high $n$ for an object with an exponential disk but a central cusp in
the light profile (eg, due to an AGN, cluster, foreground star, noise, etc).

Galaxies from the three COSMOS surveys all have morphologies that are
reasonably well-predicted by the GZH based on \fbest. The relation is strongest
for early-type galaxies, since the comparable fractions of irregulars and
spirals shows that clumpy/asymmetric morphologies are clearly also being
captured by the \fbest{} parameter. The small dropoff in the fraction of
early-types for the Cassata and ZEST samples at \fbest$<0.1$ is somewhat
puzzling, but could potentially be due to the presence of a small
population of extremely compact objects (stars or quasars) that would be
classified as artifacts in GZH, explaining the dip in \fbest.

We also emphasize that the various COSMOS catalogs are not consistent; for
example, only 64\% of the galaxies in \citet{tas11} have the same morphological
class as \citet{cas07}. The bulk of these are galaxies classified as ``spiral''
using one technique and ``irregular'' using another (\citealt{cas07} is more
likely to classify galaxies as irregular).  A detailed comparison of the
external catalogs is outside the scope of this paper, but we emphasize only
that it is \emph{impossible} for any catalog to be consistent with all three.
We interpret the broad agreements between early-types, spirals, and irregulars
for all three samples as a validation of the crowdsourced visual
classifications provided by GZH.

\begin{figure*}
\center
\includegraphics[width=1.0\textwidth]{figures/comparisons_odd.pdf}
\caption{Same as Figure~\ref{fig:comparisons_features}, but comparing external
morphologies to \fodd{} in GZH. Measurement of \fodd{} excludes galaxies
dominated by lenses or dust lanes ($<1\%$ of the total), which are GZH
categories not expected to correlate with larger-scale structural parameters.}
\label{fig:comparisons_odd}
\end{figure*}

Figure~\ref{fig:comparisons_odd} analyzes the same sets of matched galaxies, but
compares the data the \fodd{} parameter in GZH to examine its correlation with
irregular morphologies. The relationship in this case is less clear; for the
COSMOS catalogs, in particular, early-types, spirals, and irregulars are
roughly equally probable at \fodd$>0.5$. In both AEGIS and GOODS (there is no
measurment of irregular structure for GEMS in \citealt{hau07}), the proportion
of irregular galaxies does increase with \fodd. However, the actual number of
galaxies at \fodd$>0.5$ is quite small ($N\lesssim100$) in both samples, and so
estimating an accurate fraction is less precise.  We also note that this
particular question ({\it ``anything odd?''}) has somewhat uncertain phrasing
in GZH, and that the design of the tree explicitly separates clumpy from
irregular or merging morphologies, which would have similar properties in
automatic techniques. A more accurate selection of irregular morphologies in
GZH should likely use some combination of \fbest, \fclumpy, and \fodd.

We repeated all of the analysis comparing morphologies between GZH and external
catalogs on the entire catalog, rather than the volume-limited sample
($z<1.0,m<22.5$) discussed above. If all redshifts and the original magnitude
limit of $m<23.5$ are used, the correlations between all of the automatic
morphological populations and GZH decrease significantly.  The only exception
are the GEMS galaxies, which are the only sample visually classified (similar
to GZH) and had a shallower magnitude cut on their catalog. Based on these
comparisons, we suggest that use of the GZH catalog be limited to galaxies with
$m<22.5$; fainter targets at the sensitivity of the \hst{} Legacy surveys have
been demonstrated to be less reliable, regardless of the method of
morphological classification being used.

\subsection{Galaxy Zoo: Hubble and Galaxy Zoo: CANDELS}\label{ssec:gzh_gzc}

Finally, there are 7,681~galaxies in the GOODS-S field with morphological
classifications in both GZH and the Galaxy~Zoo:~CANDELS project (Simmons
et~al., submitted). Since both the sensitivity and filters for the two sets of
images are significantly different (and there is no correction for redshift
bias applied to GZC), there is no prior reason to expect a perfect correlation
between the separate vote fractions for the projects. Briefly, we note that the
\ffeatures{} value for GZH is on average higher than GZC; the effect is
strongest at $f_\mathrm{features,GZC}<0.3$, for which roughly half the galaxies
have $f_\mathrm{features,GZH}>0.5$. However, the correlation between vote
fractions is single-valued (although not linear, with a Pearson correlation
coefficient of $r=0.6$), and should be possible to calibrate using a similar
approach to that described in Section~\ref{sec:debiasing}; the correlation
between other tasks, such as edge-on galaxies is significantly stronger
($r=0.9$). So while the raw vote fractions are not directly comparable, the
initial analysis indicates that the broad morphologies are at least consistent.
